Clustering in a hyperbolic model of complex networks.

### 2020-01-02 clustering_infinite_model.tex
version 1.6

As mentioned earlier, we plan to make use of the Campbell-Mecke formula for comparing the clustering coefficient and function of $\GPo$ with certain quantities associated with $\Ginf$. We will be considering the Poisson process $\Pcal$ to which we add one additional point $(0,y)$ on the $y$-axis. In some computations the height $y$ will be fixed, but eventually we shall take it exponentially distributed with parameter $\alpha$, and independent of $\Pcal$.
We refer to $(0,y)$ as ``the typical point''.

To provide some intuition for this definition and name, note that we can alternatively view $\Pcal$ as follows. 
We take a constant intensity Poisson process on $\eR$ corresponding to the $x$-coordinates, and to each point
we attach a random ``mark'', corresponding to the $y$-coordinate, where the marks are i.i.d.~exponentially distributed with parameter $\alpha$.

@@

\subsection{The degree of the typical point\label{ssec:degrees_infinite_model}}

Before considering clustering we briefly investigate the distribution of the degree of the typical point. 
For $p = (x,y) \in \eR\times[0,\infty)$ we define
\begin{equation}\label{eq:def_degree_inf_model}
	\rho(p,k) := \Prob{\Po(\Mu{p}) = k},
\end{equation}
where $\Po(\lambda)$ denotes a Poisson random variable with expectation $\lambda$. We will often write $\rho(y,k)$ instead of $\rho(p,k)$.

Let the random variable $D$ denote the degree of the typical point. Since the typical point has a height that is independent of the Poisson process and exponential($\alpha$)-distributed, for $k \in \N_0$:
\begin{equation}\label{eq:def_pk}
	p_k :=  \Prob{D = k} = \int_0^\infty \rho(y,k) \alpha e^{-\alpha y} \, dy.
\end{equation}

Using the transformation of variables $z = \xi e^{\frac{y}{2}}$ (so $dy = \frac{2}{z}dz$), we compute
\begin{align*}
	p_k
    &= \frac{1}{k!} \int_0^\infty \left(\xi e^{\frac{y}{2}}\right)^k 
    	e^{-\xi e^{\frac{y}{2}}} \alpha e^{-\alpha y} \, dy
    	= \frac{\alpha \xi^{2\alpha}}{k!} \int_0^\infty 
    	\left(\xi e^{\frac{y}{2}}\right)^{k - 2\alpha} e^{-\xi e^{\frac{y}{2}}}
        \, dy\\
    &= \frac{2\alpha \xi^{2\alpha}}{k!} \int_{\xi}^{\infty} 
    	z^{k -2\alpha-1} e^{-z} \, dz
    	= \frac{2\alpha \xi^{2\alpha}\Gamma^+(k - 2\alpha, \xi)}{k!},
\end{align*}
where we recall that $\Gamma$ denotes the Gamma-function and $\Gamma^{+}$ the upper incomplete Gamma-function.
Note that, unsurprisingly, this is identical to the expression Gugelmann et al.~\cite{gugelmann2012random} gave for the limiting degree distribution of $G(n;\alpha,\nu)$. Using Stirling's approximation to the gamma function, we find that

\begin{equation}\label{eq:degree_distribution_P_asymptotics}
	p_k \sim 2\alpha\xi^{2\alpha} k^{-(2\alpha + 1)}
	\quad \text{as } k \to \infty.
\end{equation}

By a similar computation we have the following result, which will be useful later on. For any $\beta > 0$, as $k \to \infty$
\begin{equation}\label{eq:general_integral_rho_y_k}
	\int_0^\infty e^{\beta y} \rho(y, k) \alpha e^{-\alpha y} \, dy
    \sim 2\alpha \xi^{2(\beta + \alpha)} k^{-2(\beta + \alpha)-1}.
\end{equation}

@@

With the asymptotic behavior of $P(y)$ we are almost ready to prove Proposition~\ref{prop:asymp}. First we will prove a result that will allow us to limit the values of $y$, when performing the integration. For this, fix some $C > 0$ and define
\[
	a^\pm(k) = 2 \log\left(\frac{k \pm C \sqrt{k \log(k)}}{\xi} \vee 1\right).
\] 
We will show that, as $k \to \infty$,
\begin{equation}\label{eq:asymptotics_average_clustering_concentration}
	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
	= (1+\smallO{1}) \int_{a^-(k)}^{a^+(k)} P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y.
\end{equation}

To establish~\eqref{eq:asymptotics_average_clustering_concentration} recall that $\mu(y) = \xi e^{\frac{y}{2}}$ and consider $\rho(y,k) = \Prob{\Po(\mu(y)) = k}$ as a function of $y$. Then, since $\mu^\prime(y) = \mu(y)/2$, we get that
\[
	\frac{\partial \rho(y,k)}{\partial y} = \frac{1}{2}\left(k - \mu(y)\right)\rho(y,k),
\]
which implies that $\rho(y,k)$ attains its maximum at $\mu(y) = k$. Moreover we see that the derivative is strictly positive when $\mu(y) < k$ and strictly negative when $\mu(y) > k$. Since $\mu(a^-(k)) < k$ and $\mu(a^+(k)) > k$, we conclude that $\rho(y,k)$, as a function of $y$, is strictly increasing on $[0,a^-(k)]$ and strictly decreasing on $[a^+(k),\infty)$. Hence, using that $P(y) \le 1$,
\begin{align*}
	&\hspace{-20pt}\int_{\R_+ \setminus [a^-(k), a^+(k)]} P(y) \rho(y,k) 
		\alpha e^{-\alpha y} \dd y\\
    &\le \int_0^{a^-(k)} \rho(y,k) \alpha e^{-\alpha y} \dd y 
    	+ \int_{a^+(k)}^{\infty}\rho(y,k) \alpha e^{-\alpha y}x \dd y \\
   	&\le \rho(a^-(k),k)\int_0^{a^-(k)} e^{-\alpha y} \, \dd y
   		+ \rho(a^+(k),k) \int_{a^+(k)}^{\infty} e^{-\alpha y} \, \dd y\\
   	&= \bigO{1}\left(\rho(a^-(k),k) + \rho(a^+(k),k)\right),
\end{align*}
as $k \to \infty$. Next we show that
\[
	\rho(a^\pm(k),k) = \bigO{k^{-(1+C^2)/2}}.
\]
Since the arguments are almost completely identical, we give the prove for $a^+(k)$.

Using Stirling's formula $k! \sim \sqrt{2\pi} k^{k + 1/2} e^{-k}$ as $k \to \infty$ we write
\begin{align*}
	\rho(a^+(k),k) &= \frac{\mu(a^+(k))^{k}}{k!} e^{-\mu(a^+(k))} \\
	&\sim (2\pi)^{-1/2} k^{-1/2} \left(\frac{\mu(a^+(k))}{k}\right)^{k} e^{-(\mu(a^+(k) - k)}\\
	&= (2\pi)^{-1/2} k^{-1/2} 
		e^{-k\left(\frac{\mu(a^+(k))}{k} - 1 - \log\left(\frac{\mu(a^+(k))}{k}\right)\right)}.
\end{align*}
Since, for sufficiently large $k$,
\[
	\frac{\mu(a^+(k))}{k} = 1 + C \sqrt{\frac{\log(k)}{k}}
\]
and $x - \log(1 + x) \sim x^2/2$ as $x \to 0$, we get 
\begin{align*}
	\rho(a^+(k),k) 
	&\sim \sqrt{2\pi} k^{-1/2} 
		e^{-k \left(C \sqrt{\frac{\log(k)}{k}} - \log\left(1 + C \sqrt{\frac{\log(k)}{k}}\right)\right)}\\
	&\sim (2\pi)^{-1/2} k^{-1/2} e^{-\frac{k \left(C \sqrt{\frac{\log(k)}{k}}\right)^2}{2}} \\
	&= \bigO{k^{-(1+C^2)/2}}.
\end{align*}

Since $C > 0$ can be chosen arbitrarily large we conclude that
\[
	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
	= (1+\smallO{1}) \int_{a^-(k)}^{a^+(k)} P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y,
\]
as $k \to \infty$. Note that this implies that if $P(y) = h(y)(1 + \smallO{1})$, uniformly in $y$, as $y \to \infty$, then 
\begin{equation}\label{eq:asymptotics_average_clustering_error_term}
	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
	\sim \int_0^\infty h(y) \rho(y,k) \alpha e^{-\alpha y} \dd y,
\end{equation}
as $y \to \infty$.

>>>>

Similar to the proof for the degrees we plan to make use of the Campbell-Mecke formula for comparing the clustering coefficient and function of $\GPo$ with certain quantities associated with $\Ginf$. Recall that a \emph{typical point} is a point $p_0 = (0,y)$, where in some computations the height $y$ will be fixed, but eventually we shall take it exponentially distributed with parameter $\alpha$, and independent of $\Pcal$.

@@

@@

With the asymptotic behavior of $P(y)$ we are almost ready to prove Proposition~\ref{prop:asymp}. First we will prove a result that will allow us to limit the values of $y$, when performing the integration. For this, recall that for any $C > 0$ we defined
\[
	y_{k,C}^\pm = 2 \log\left(\frac{k \pm C \sqrt{k \log(k)}}{\xi} \vee 1\right).
\] 
Since $P(y) \le 1$ by the concentration of heights results (Proposition~\ref{prop:concentration_height_general}) we have that
We will show that, as $k \to \infty$,
\begin{equation}\label{eq:asymptotics_average_clustering_concentration}
	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
	= (1+\smallO{1}) \int_{y_{k,C}^-}^{y_{k,C}^+} P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y.
\end{equation}

Note that this implies that if $P(y) = h(y)(1 + \smallO{1})$, uniformly in $y$, as $y \to \infty$, then 
\begin{equation}\label{eq:asymptotics_average_clustering_error_term}
	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
	\sim \int_0^\infty h(y) \rho(y,k) \alpha e^{-\alpha y} \dd y,
\end{equation}
as $y \to \infty$.

%To establish~\eqref{eq:asymptotics_average_clustering_concentration} recall that $\mu(y) = \xi e^{\frac{y}{2}}$ and consider $\rho(y,k) = \Prob{\Po(\mu(y)) = k}$ as a function of $y$. Then, since $\mu^\prime(y) = \mu(y)/2$, we get that
%\[
%	\frac{\partial \rho(y,k)}{\partial y} = \frac{1}{2}\left(k - \mu(y)\right)\rho(y,k),
%\]
%which implies that $\rho(y,k)$ attains its maximum at $\mu(y) = k$. Moreover we see that the derivative is strictly positive when $\mu(y) < k$ and strictly negative when $\mu(y) > k$. Since $\mu(a^-(k)) < k$ and $\mu(a^+(k)) > k$, we conclude that $\rho(y,k)$, as a function of $y$, is strictly increasing on $[0,a^-(k)]$ and strictly decreasing on $[a^+(k),\infty)$. Hence, using that $P(y) \le 1$,
%\begin{align*}
%	&\hspace{-20pt}\int_{\R_+ \setminus [a^-(k), a^+(k)]} P(y) \rho(y,k) 
%		\alpha e^{-\alpha y} \dd y\\
%    &\le \int_0^{a^-(k)} \rho(y,k) \alpha e^{-\alpha y} \dd y 
%    	+ \int_{a^+(k)}^{\infty}\rho(y,k) \alpha e^{-\alpha y}x \dd y \\
%   	&\le \rho(a^-(k),k)\int_0^{a^-(k)} e^{-\alpha y} \, \dd y
%   		+ \rho(a^+(k),k) \int_{a^+(k)}^{\infty} e^{-\alpha y} \, \dd y\\
%   	&= \bigO{1}\left(\rho(a^-(k),k) + \rho(a^+(k),k)\right),
%\end{align*}
%as $k \to \infty$. Next we show that
%\[
%	\rho(a^\pm(k),k) = \bigO{k^{-(1+C^2)/2}}.
%\]
%Since the arguments are almost completely identical, we give the prove for $a^+(k)$.
%
%Using Stirling's formula $k! \sim \sqrt{2\pi} k^{k + 1/2} e^{-k}$ as $k \to \infty$ we write
%\begin{align*}
%	\rho(a^+(k),k) &= \frac{\mu(a^+(k))^{k}}{k!} e^{-\mu(a^+(k))} \\
%	&\sim (2\pi)^{-1/2} k^{-1/2} \left(\frac{\mu(a^+(k))}{k}\right)^{k} e^{-(\mu(a^+(k) - k)}\\
%	&= (2\pi)^{-1/2} k^{-1/2} 
%		e^{-k\left(\frac{\mu(a^+(k))}{k} - 1 - \log\left(\frac{\mu(a^+(k))}{k}\right)\right)}.
%\end{align*}
%Since, for sufficiently large $k$,
%\[
%	\frac{\mu(a^+(k))}{k} = 1 + C \sqrt{\frac{\log(k)}{k}}
%\]
%and $x - \log(1 + x) \sim x^2/2$ as $x \to 0$, we get 
%\begin{align*}
%	\rho(a^+(k),k) 
%	&\sim \sqrt{2\pi} k^{-1/2} 
%		e^{-k \left(C \sqrt{\frac{\log(k)}{k}} - \log\left(1 + C \sqrt{\frac{\log(k)}{k}}\right)\right)}\\
%	&\sim (2\pi)^{-1/2} k^{-1/2} e^{-\frac{k \left(C \sqrt{\frac{\log(k)}{k}}\right)^2}{2}} \\
%	&= \bigO{k^{-(1+C^2)/2}}.
%\end{align*}
%
%Since $C > 0$ can be chosen arbitrarily large we conclude that
%\[
%	\int_0^\infty P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y
%	= (1+\smallO{1}) \int_{a^-(k)}^{a^+(k)} P(y) \rho(y,k) \alpha e^{-\alpha y} \dd y,
%\]
%as $k \to \infty$.
